---
title: "Project Prosoal - victoria-public-transport-visualisation "
author: "Wei Wang"
format: html
editor: visual
---

## 1. Load and View the general information of dataset

```{r}
install.packages("skimr")
pacman::p_load(skimr, dplyr, tidyverse)
roadwork<- read.csv("~/Downloads/data visualization /datasets/monthly_public_transport_patronage_by_mode.csv")
roadwork
skim(roadwork)
glimpse(roadwork)
```

#### Summary of all variables

```{r}
summary(roadwork)
```

#### Make sure Regional.Bus is numeric

```{r}
library(tidyverse)
roadwork <- roadwork %>%
  mutate(
    Regional.bus = as.numeric(Regional.bus)
  )
```

#### Find the NA value

```{r}
sum(is.na(roadwork$Regional.bus))
```

```{r}
which(is.na(roadwork$Regional.bus))
roadwork[83, ]
roadwork %>%
  slice(83) %>%
  select(Year, Month, Month.name, Regional.bus, Total_usage)
```

#### Create a new column "Total_usage" each year for each mode

```{r}
roadwork %>%
  mutate(
    Total_usage = ( Metropolitan.train + Metropolitan.tram + Metropolitan.bus + Regional.bus + V.Line.coach + V.Line.train
    ) 
    )

```

#### Find the month/year with the highest and lowest total usage

```{r}
roadwork %>%
  filter(Total_usage == max(Total_usage, na.rm = TRUE)) %>%
  select(Month.name, Year, Total_usage)

roadwork %>%
  filter(Total_usage == min(Total_usage, na.rm = TRUE)) %>%
  select(Month.name, Year, Total_usage)
```

**Graph1: Annual Total Patronage By year**

```{r}
library(ggplot2)
annual_usage <- roadwork %>%
  group_by(Year) %>%
  summarise(
    total_usage = sum(Total_usage, na.rm = TRUE)
  )

ggplot(annual_usage, aes(x = factor(Year), y = total_usage)) +
  geom_col(fill = "#2C7FB8", width = 0.65) +
  labs(
    title = "Annual Public Transport Patronage in Victoria",
    x = "Year",
    y = "Total annual patronage"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
  
```

#### Graph 2: Monthly Patronage Trends by Transport Mode

```{r}
ggplot(roadwork_long, aes(x = Date, y = Patronage, colour = Mode)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Monthly Public Transport Patronage by Mode in Victoria",
    x = "Year",
    y = "Monthly patronage",
    colour = "Transport mode"
  ) +
  theme_minimal()
```

#### Graph 2.1: Create a separate small chart for each Mode

```{r}
ggplot(roadwork_long, aes(x = Date, y = Patronage, colour = Mode)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  facet_wrap(~ Mode, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Monthly Public Transport Patronage Trends by Mode",
    x = "Year",
    y = "Monthly patronage"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

#### Graph 3: Total Patronage by Transport Mode

```{r}
mode_totals <- roadwork %>%
  summarise(
    Metropolitan_train = sum(Metropolitan.train, na.rm = TRUE),
    Metropolitan_tram = sum(Metropolitan.tram, na.rm = TRUE),
    Metropolitan_bus = sum(Metropolitan.bus, na.rm = TRUE),
    V_Line_train = sum(V.Line.train, na.rm = TRUE),
    V_Line_coach = sum(V.Line.coach, na.rm = TRUE),
    Regional_bus = sum(Regional.bus, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Mode",
    values_to = "Total_patronage"
  )

ggplot(mode_totals, aes(x = reorder(Mode, Total_patronage), y = Total_patronage)) +
  geom_col(fill = "#41B6C4") +
  coord_flip() +
  labs(
    title = "Total Patronage by Transport Mode",
    x = "Transport mode",
    y = "Total patronage"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

**Graph 4: Average Patronage by Season**

```{r}
roadwork$Season <- factor(
  roadwork$Season,
  levels = c("Spring", "Summer", "Autumn", "Winter")
)
levels(roadwork$Season)
ggplot(roadwork, aes(x = Season, y = Total_usage, fill = Season)) +
  geom_boxplot(alpha = 0.8, width = 0.6) +
  labs(
    title = "Monthly Public Transport Patronage by Season",
    x = "Season",
    y = "Monthly patronage"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```
